### 2014-04-20 patchy-selection-paper.tex
.

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are given by a spatial branching process.
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\partial B(r_0)$ its boundary.
Suppose that there is a constant outflux of migrant families of rate $\gamma(x)$ from each point $x \in \partial B(r_0)$.
To make expressions \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate} more concrete,
we need to replace ``outflux of families'' by a sum over points at distance $r_0$ from the patch
of the outflux $\gamma$ at that point, mulitplied by the relevant second quantity.
By examining the constituent terms,
it will turn out that \eqref{eqn:gestalt_migrate} is a constant multiple of \eqref{eqn:gestalt_q}.

@@

Now we turn to the expression \eqref{eqn:gestalt_migrate} for the outflux of successful migrants.
Consider the probability that a migrant family beginning with a single individual at $x$ will ever establish in the new patch,
which we assume is circular, has relatively small area $A$ and is located near position $y$.
% The probability that this family reaches $A$ at time $t$ is the probability that it survives until $t$, $(1-p_e(t))$,
% multiplied by the probability that the family reaches the patch, given survival.
% If $t$ is large then as in the previous section, $(1-p_e(t)) \approx e^{-\alpha t}/\E[K]$.
It would be possible to analyze directly the probability that a given migrant family establishes in $A$,
as in \citet{barton1987probability,branchingbrownian};
but we make a simpler approximation, approximating this by the chance that the ``trunk'' hits the new patch,
multiplied by the chance that at least member of the family escapes demographic stochasticity to establish.
Write $h(x,y)$ for the probability that the Brownian trunk
hits the new patch centered at $y$ before the family dies out,
and $f(A)$ for the chance that the family manages to establish in the new patch,
given that it successfully arrives.
As for $q(y)$ above, expression \eqref{eqn:gestalt_migrate} is an integral of $h(x,y)$:
\begin{align}\label{eqn:migrate_integral}
  \migrate(y) &= \int_{\partial B(r_0)} \gamma(x) h(x,y) f(A) dx .
\end{align}

Intuitively, $h(x,y)$ counts each family once if it hits the patch,
while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
Since the expected total time spent in the patch does not depend on the distance at which the family started
if we know that it survived to hit the patch,
it is intuitively reasonable that $u(x,y)$ and $h(x,y)$ are simply related.
Using this fact,
and that $(1-p_e(t)) \approx e^{-\alpha t}/\E[K]$,
it turns out (see Appendix~\ref{apx:hitting_occupation}) that
\begin{align} 
  h(x,y) = \frac{ s(A) }{ \E[K] } u(x,y),
\end{align}
where $s(A)$ is the total expected occupation time of the patch given the family arrives,
and recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \ref{eqn:occupation_integral} and \ref{eqn:migrate_integral} only differ by a constant factor
(at least asymptotically for large distances),
we now know that the outflux of successful migrant families is equal to the mean occupation frequency
multiplied by $s(A) f(A) / \E[K]$.
The probability $f(A)$ that the family does not die out once it arrives at the new patch
is approximately $\E[1-(1-p_e)^K] \approx 2 s_p \E[K] / \xi^2$.
% If we let $w$ be the width of the patch, then the remaining term is
% \begin{align}
%   s(w) \simeq \frac{1}{2\alpha} \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
% \end{align}
Combining these terms with the expression for $s(A)$ from Appendix~\ref{apx:hitting_occupation},
\begin{align}
  \migrate(y) \approx q(y) \times \frac{ s_p  }{ \alpha \xi^2 } 
  \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
\end{align}

@@

Combining this with \eqref{eqn:bp_prob_estab}, as in \eqref{eqn:gestalt_q}, 
the rate of establishment, for a patch of area $A$ at distance $r$ from an already occupied patch, is approximately
\begin{align} \label{eqn:migrate}
    \migrate(r) = \gamma F(r) = \frac{\rho \sigma s_p\sqrt{2s_m} }{\xi^2}  \exp\left( -\frac{ \sqrt{2 s_m} r}{\sigma} \right).
\end{align}
Note that the exponential term is expected based on \eqref{eqn:eqfreq} -- 
since establishment requires rare long distant migrants to be present, and \eqref{eqn:eqfreq} gives the density of such --
but the first term required other arguments.

@@

We will assume that once a $B$ allele is introduced (by migration or mutation), 
it becomes established in the poor habitat patch rapidly. 
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the the second patch adapts
simply comes down to whether the a new mutation or a migrant allele is the first to become established in the second patch.

Taking expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above,
we see that if the patches each have sufficiently large area,
the rate of migration and mutation will be on the same order if 
$\mu \approx ((s_p+s_m)/2) \exp(-\sqrt{2 s_m} r / \sigma)$,
\plr{check}
where $r$ is the distance between the patches.
(Note that in both cases, $A$ is the area of the not-yet-adapted patch.)

If we take $\mu = 10^{-8}$, and $s_p=s_m=s$, then this is satisfied if $r/\sigma \approx (18+\log(s))/\sqrt{s}$,
so if the selective pressure is fairly strong (say, $s=.05$),
then the distance between patches must be about 70 times the dispersal distance (i.e.\ $r/\sigma \approx 68$);
while if $s$ is much smaller, say $s = .001$, 
then the patches must be separated by many hundreds of times the dispersal distance (i.e.\ $r/\sigma \approx 368$).
The situation does not change much if $\mu$ is larger -- even if $\mu = 10^{-6}$, 
the required separation between patches is only reduced by $5/\sqrt{s_m}$.
If the separation between patches is much larger than this value -- certainly, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will occur independent between patches.
On the other hand, if the separation is much smaller, then a single
mutation is likely to fix in both patches.

One way to visualize this is by a phase diagram.
In figure \ref{fig:phase_diagram},
we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
as a function of population density, mutation rate, and distance between patches (scaled by $\sqrt{s_m}/\sigma$).

\begin{figure}[ht]
  \begin{center}
    \includegraphics{phase-diagram-log}
  \end{center}
  \caption{
  Mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
  as a function of population density ($\rho$), mutation rate ($\mu$), 
  and distance to an already adapted patch ($R$), scaled by $\sqrt{s_m}/\sigma$.
  Note that population density and mutation rate are shown on a log scale.
  \plr{fix axis labels}
  \label{fig:phase_diagram}
  }
\end{figure}

If we are willing to take these approximations at face value, 
and assuming both rates are small, 
the time until the first of the two patches is colonized by the $B$ allele will be approximately exponentially distributed with mean $1/(2 \mutrate)$.
Following this, the time till the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with mean $1/(\mutrate+\migrate)$.
Both these times scale linearly with $\xi^2/(\rho A s_{B})$, \plr{check}
so that the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with patch population size and benefical selection coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adapatation, is
\begin{equation}
  \frac{\mutrate}{\mutrate+\migrate} = \frac{2\mu/(s_p+s_m)}{2\mu/(s_p+s_m) + \exp\left(- \sqrt{2 s_m} r / \sigma \right) },
\end{equation}
\plr{check}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches, at rate $\sqrt{s_m} /\sigma$.

@@

If not yet adapted,
a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$.
Without loss of generality, assume that patches $1, \ldots, k$ are already adapted,
and these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch, 
then the total rate of adaptation through migration is
\begin{equation}
  \frac{ s_p (s_p+s_m) \rho A}{\xi^2} \sum_{i=1}^{k} \exp\left(- \sqrt{2 s_m} r_i/\sigma\right),
\end{equation}
and the probability that the migrant, adapted allele 
comes from patch $i$ is $\exp\left(- \sqrt{2 s_m} r_i/\sigma\right)/\sum_{j=1}^{k} \exp\left(- \sqrt{2 s_m} r_j/\sigma\right)$.
These specify the transition rates of the process.

@@

\subsection{Temporary}

>>>>

We are approximating the dynamics of the focal allele in the region further away than $r_0$ from the patch
as the sum of independent migrant families, each of whose dynamics are given by a spatial branching process.
Call $B(r_0)$ the region closer than $r_0$ to the patch, and $\partial B(r_0)$ its boundary.
Suppose that there is a constant outflux of migrant families from each point $x \in \partial B(r_0)$,
such that the long-term proportion of individuals near a point $x$ in $\partial B(r_0)$ that initiate new migrant families
is $\gamma(x)$.
To make expressions \eqref{eqn:gestalt_q} and \eqref{eqn:gestalt_migrate} more concrete,
we need to replace ``outflux of families'' by a sum over points at distance $r_0$ from the patch
of the outflux $\gamma$ at that point, mulitplied by the relevant second quantity.
By examining the constituent terms,
it will turn out that \eqref{eqn:gestalt_migrate} is a constant multiple of \eqref{eqn:gestalt_q}.

@@

Now we turn to the expression \eqref{eqn:gestalt_migrate} for the outflux of successful migrants.
Consider the probability that a migrant family beginning with a single individual at $x$ will ever establish in the new patch,
which we assume is circular, has relatively small area $A$ and is located near position $y$.
% The probability that this family reaches $A$ at time $t$ is the probability that it survives until $t$, $(1-p_e(t))$,
% multiplied by the probability that the family reaches the patch, given survival.
% If $t$ is large then as in the previous section, $(1-p_e(t)) \approx e^{-\alpha t}/\E[K]$.
It would be possible to analyze directly the probability that a given migrant family establishes in $A$,
as in \citet{barton1987probability,branchingbrownian};
but we make a simpler approximation, approximating this by the chance that the ``trunk'' hits the new patch,
multiplied by the chance that at least member of the family escapes demographic stochasticity to establish.
Write $h(x,y)$ for the probability that the Brownian trunk
hits the new patch centered at $y$ before the family dies out,
and $f(A)$ for the chance that the family manages to establish in the new patch,
given that it successfully arrives.
As for $q(y)$ above, expression \eqref{eqn:gestalt_migrate} is an integral of $h(x,y)$:
\begin{align}\label{eqn:migrate_integral}
  \migrate(y) &= \rho \int_{\partial B(r_0)} \gamma(x) h(x,y) f(A) dx .
\end{align}

Intuitively, $h(x,y)$ counts each family once if it hits the patch,
while the occupation density $u(x,y)$ counts the total amount of time that the family spends in the patch.
Since the expected total time spent in the patch does not depend on the distance at which the family started
if we know that it survived to hit the patch,
it is intuitively reasonable that $u(x,y)$ and $h(x,y)$ are simply related.
Using this fact,
and that $(1-p_e(t)) \approx e^{-\alpha t}/\E[K]$,
it turns out (see Appendix~\ref{apx:hitting_occupation}) that
\begin{align} 
  h(x,y) = \frac{ s(A) }{ \E[K] } u(x,y),
\end{align}
where $s(A)$ is the total expected occupation time of the patch given the family arrives,
and recall that $\E[K]$ is the expected family size after a large number of generations, given survival.
Since the integrands in expressions \ref{eqn:occupation_integral} and \ref{eqn:migrate_integral} only differ by a constant factor
(at least asymptotically for large distances),
we now know that the outflux of successful migrant families is equal to the mean occupation frequency
multiplied by $s(A) f(A) / \E[K]$.
The probability $f(A)$ that the family does not die out once it arrives at the new patch
is approximately $\E[1-(1-p_e)^K] \approx 2 s_p \E[K] / \xi^2$.
% If we let $w$ be the width of the patch, then the remaining term is
% \begin{align}
%   s(w) \simeq \frac{1}{2\alpha} \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
% \end{align}
Combining these terms with the expression for $s(A)$ from Appendix~\ref{apx:hitting_occupation},
\begin{align}
  \label{eqn:migrate}
  \migrate(y) \approx q(y) \times \rho \frac{ s_p  }{ \alpha \xi^2 } 
  % \left( 1 - \left( \frac{\pi w \sqrt{2 \alpha}}{ 2 \sigma } \right)^{\frac{d-1}{2}} e^{-w \sqrt{2\alpha}/\sigma} \right).
\end{align}
For patches that are small relative to $\sigma/\sqrt{2s_m}$, there is an additional correction term (see Appendix~\ref{apx:hitting_occupation}),
but at this point we would be violating various other assumptions above, so we omit it.

@@

Combining this with \eqref{eqn:bp_prob_estab}, as in \eqref{eqn:gestalt_q}, 
the rate of establishment, for a patch of area $A$ at distance $r$ from an already occupied patch, is approximately
\begin{align} % \label{eqn:migrate}
    \migrate(r) = \gamma F(r) = \frac{\rho \sigma s_p\sqrt{2s_m} }{\xi^2}  \exp\left( -\frac{ \sqrt{2 s_m} r}{\sigma} \right).
\end{align}
Note that the exponential term is expected based on \eqref{eqn:eqfreq} -- 
since establishment requires rare long distant migrants to be present, and \eqref{eqn:eqfreq} gives the density of such --
but the first term required other arguments.

@@

We will assume that once a $B$ allele is introduced (by migration or mutation), 
it becomes established in the poor habitat patch rapidly. 
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch.

Taking expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above,
we see that if the patches each have sufficiently large area,
the rate of migration and mutation will be on the same order if 
$A \mu \approx (1/s_m) \exp(- R \sqrt{2 s_m} / \sigma)$,
where $R$ is the distance between the patches,
and recall that $A$ is the area of the not-yet-adapted patch.
In other words, the migrational analogue of ``mutational influx'' ($\mu \rho A$) 
is $\rho (1/s_m) \exp(- R \sqrt{2 s_m} / \sigma)$,
which depends fairly strongly on the selective disadvantage $s_m$ between patches.
Equivalently, migration will occur much quicker than mutation 
if $R/\sigma > (\log(1/s_m) - \log(A\mu))/\sqrt{2s_m}$.

If we take the mutational influx in a patch to be one per 1,000 generations ($A \mu = .001$),
and the selection against transiting alleles $s_m = .01$,
adaptation procedes independently for distances $R/\sigma > \log(s_m)/\sqrt{s_m}$,
so if the selective pressure is fairly strong (say, $s_m=.05$),
then the distance between patches must be about 30 times the dispersal distance (i.e.\ $R/\sigma > 30$),
while if $s_m$ is much smaller, say $s_m = .001$, 
then the patches must be separated by several hundred times the dispersal distance (i.e.\ $R/\sigma > 300$).
The situation does not change much if the influx is quite a bit higher -- even if $A \mu = 1$,
the required separation between patches is only reduced by $7/\sqrt{s_m}$,
to $R/\sigma > 10$ with $s_m=.05$ and $R/\sigma>150$ for $s_m=.001$.
If the separation between patches is much larger than this transitional value -- certainly, by a factor of ten -- 
then patches are effectively isolated -- most adaptation will occur independent between patches.
On the other hand, if the separation is much smaller, then a single
mutation is likely to fix in both patches.

One way to visualize this is by a phase diagram.
In figure \ref{fig:phase_diagram},
we give a contour plot for the time until adaptation by migration ($\Tmig = 1/\migrate$) and by new mutation ($\Tmut=1/\mutrate$)
as a function of various parameters.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{phase-diagram-log}
  \end{center}
  \caption{
  Shaded contour plots of
  mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
  as a function of population density ($\rho$), 
  distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
  strength of deleterious selection ($s_m$),
  and mutation rate scaled by patch area ($\mu A$).
  The line at which $\Tmig=\Tmut$ is shown in black.
  Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=50$, and $\xi^2=1$.
  Note that population density and mutation rate are shown on a log scale.
  \label{fig:phase_diagram}
  }
\end{figure}

If we are willing to take these approximations at face value, 
and assuming both rates are small, 
the time until the first of the two patches is colonized by the $B$ allele will be approximately exponentially distributed with mean $1/(2 \mutrate)$.
Following this, the time till the second patch is subsequently colonized 
(via either migration or new mutation) 
will be exponentially distributed with mean $1/(\mutrate+\migrate)$.
Both these times scale linearly with $\xi^2/(\rho s_{B})$, 
so that the overall time to adaptation is robustly predicted to increase with offspring variance ($\xi^2$)
and decrease with population density and benefical selection coefficient.
Furthermore, the probability that the second adaptation is a new mutation,
i.e.\ the probability of parallel adapatation, is in $d=1$ ($d=2$ only differs slightly)
\begin{equation}
  \frac{\mutrate}{\mutrate+\migrate} = \frac{2 A \mu s_m}{2 A \mu s_m + \exp\left(- \sqrt{2 s_m} r / \sigma \right) },
\end{equation}
so that the probability of parallel mutation should increase
approximately logistically with the distance between the patches, at rate $\sqrt{s_m} /\sigma$.

@@

If not yet adapted,
a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$.
Without loss of generality, assume that patches $1, \ldots, k$ are already adapted,
and these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch, 
then the total rate of adaptation through migration is approximately
\begin{equation}
  \frac{ s_p \rho }{s_m \xi^2} \sum_{i=1}^{k} \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and the probability that the migrant, adapted allele 
comes from patch $i$ is 
\begin{align}
  \exp\left(- r_i \sqrt{2 s_m} /\sigma\right)/\sum_{j=1}^{k} \exp\left(- r_j \sqrt{2 s_m} /\sigma\right) .
\end{align}
These specify the transition rates of the process.

@@

\subsection{Hitting and occupation}
\label{apx:hitting_occupation}
